Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-18 doc/ApTranscriptome_manuscript.Rmd
Completed first draft manuscript sent to co-authors

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

## Abstract

- When ants are exposed to different temperatures, lots of genes change expression. 
- The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress. 
- Comparisons between species indicate that *A. carolinensis*, which has greater basal gene expression and physiological thermal tolerance, tolerates thermal stress by shutting down gene expression. *A. picea* tolerates thermal stress by activating protective genes

## Introduction

@@

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology. To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.

For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [cite..] and provide insight on evolutionary potential. However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [cite...] and miss genes with small expression changes that have large phenotypic effects [cite...], thus limiting their usefullness in this context. Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes is over-estimated while many causal genes of small effect are missed [@beavis; @larson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction. Analog

In this study, we apply a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal extremes to overcome the above limitations. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach. Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [cite...]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species.

To explore differences in the thermal reactionomes between *A. picea* and *A. carolinensis*, we performed a whole organism RNA sequencing study spanning temperatures from 0° to 38.5°C. We identified genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), compared the patterns

3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures. We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ... in the thermally-responsive group.

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. We compared the thermal reactionomes between two parapatric species of *Aphaenogaster*, *A. carolinensis* from North Carolina, USA and *A. picea* from Vermont, USA, to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance. Specifically, we examined differences between the species in (i) the number of thermally-responsive genes, (ii) the distribution of responsive genes into functional categories, (iii) basal expression near optimal temperatures for responsive genes and (iv) the thermal sensitivity of the transcriptome. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolinensis* [@diamond2013] we predicted that it would show more plasticity of gene expression in response to heat shock, and lower basal levels of expression for heat-tolerant genes.

## Methods

### Samples

@@

### Sequencing, assembly and annotation

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmeider2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

### Thermally-responsive genes

We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. In addition, we used the web-tool DAVID for ....

### Species-level comparisons

To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category *Not Responsive*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A. picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A. picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A. picea* than *A. carolinensis*.

Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$. We then partitioned the contribution of each measure by dividing it by $\Delta T$.

To test if genes that were responsive to cold temperatures in the southern *A. carolinensis* species or hot temperatures in the northern *A. picea* species had higher basal expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. carolinensis*. To test this with our data, for each **Intermediate** expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of **Bimodal** expressed genes would be greater in *A. carolinensis* than *A. picea*. We performed the same analysis as with the **Intermediate** transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity.

Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes. We then tested if the temperature at which the **High** or **Low** group differed among the two species.

### Data availability

@@

## Results

@@

### Thermally-responsive genes

@@

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. These differences were not simply of magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

### Species comparisons

@@

### Gene set enrichment analysis

@@

### Candidate genes

@@

## Discussion

@@

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 2). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig. 1, Fig. 2). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [cite...kingsolver, chown]. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal2012].

For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed  [@barshis2013]. However, in contrast to these expectations, we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both spcies. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

@@

To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodginsdavis2012]). By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Thus, we were able to compare the temperature at which genes were activated near the upper and lower thermal limits between species (Fig. 2). For genes activated at cold temperatures, there was little difference among species with *A. picea* genes turning on at 13°C compared to 12.8°C for *A. carolinensis*. For high temperature genes, *A. carolinensis* genes were turned on at 33.8°C compared to 32.9°C for *A. picea*.

### Caveats/limitations

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, would considerably resolve this problem. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change. Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...]. In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].

### Conclusions

In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A. carolinensis*, to date. Using a novel regression approach, we found that a large proportion of the transcriptome is thermally-responsive. These data indicate that the two species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock and other protective molecules, while *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace. For both species, transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures. These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

## References

>>>>

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

\linenumbers
\tableofcontents

\newpage

# Abstract

Thermal tolerance is a key trait governing the distributions of species, their abundances, and underlies their future responses to climate change. However, the genetic basis of thermal tolerance is poorly understood for most species. In this study, we present a novel regression approach to characterize the thermal reaction norm for all genes in an organism's transcriptome. We apply this approach to parapatric ant species in the eastern United States; *A. carolinensis* and *A. picea*, which have warmer and colder distributions, respectively. Using this approach, we find that at least 10% of genes are thermally responsive. In both species, more genes are up-regulated in response to cold than warm temperatures, suggesting that there is less adaptive potential to heat than cold stress. Between species, *A. carolinensis* shuts down more genes in response to thermal extremes, whereas *A. picea* activates more genes in response to cold and hot temperatures. The genes that are shut down in *A. carolinensis* but up-regulated in *A. picea* included many metabolism, cellular organization and repair gene ontology terms. We conclude that these species have different mechanisms for responding to thermal stress. *A. picea* resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. In summary, this study demonstrates how we can gain insight on the genetic basis of local adaption to climate using a reaction norm approach with transcriptomic data.

# Introduction

@@

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology. To date, RNAseq studies have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.

For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014]. However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes are over-estimated while many causal genes of small effect are missed [@beavis1994; @larsson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010]. Analogously, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient (e.g. the thermal reactionome). Instead of focusing on individual genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively.

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species. For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A. carolinensis*, or colder temperatures in *A. picea*. Additionally, genes may change their sensitivity to thermal extremes. Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance in these species. Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.

# Methods

## Samples

@@

## Sequencing, assembly and annotation

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

## Thermally-responsive genes

We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

## Species-level comparisons

We performed comparative analyses between the two species focusing on the genes that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of gene activation and sensitivity of the thermally-responsive genes. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**. To test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression for each gene in each species. We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes. We then performed a *t*-test to determine if the mean temperature of gene activation differed between the two species for each group.

To determine if the sensitivity of genes to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** genes in each species. For each **Intermediate** gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature. For the **Bimodal** genes, we performed the same analysis as with the **Intermediate** genes, but with a greater standard deviation being indicative of a more sensitive gene (convex reaction norm) while a low standard deviation being a gene with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity for that species.

To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of genes that were responsive to cold temperatures (**Low** genes) in the southern *A. carolinensis* species, or hot temperatures (**High** genes) in the northern *A. picea* species using non-parametric two-sided Wilcoxon tests.

Finally, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A. picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A. picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A. picea* than *A. carolinensis*. Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$. We then partitioned the contribution of each measure by dividing it by $\Delta T$.

## Gene set enrichment analysis

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.

## Data availability

@@

# Results

@@

## Thermally-responsive genes

@@

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. These differences were not only in magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

## Species comparisons

@@

## Gene set enrichment analysis

@@

**Table** Table of enriched GO terms for each expression category in each species.

## Candidate genes

@@

# Discussion

@@

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.

```{r max_exp_PDF, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
  geom_density(alpha=0.2, position="identity") + 
  labs(x = "Temperature of maximum expression", y = "Density")

print(maxexplot)
```

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellermann et al [-@kellermann2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed [@pigliucci2006]. However, in contrast to these expectations, we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both spcies. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

@@

To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]). By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Thus, we were able to compare the temperature at which genes were activated near the upper and lower thermal limits between species (Fig. 2). For genes activated at cold temperatures, there was little difference among species with *A. picea* genes turning on at 13°C compared to 12.8°C for *A. carolinensis*. For high temperature genes, *A. carolinensis* genes were turned on at 33.8°C compared to 32.9°C for *A. picea*.

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, would considerably resolve this problem. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change. Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@jensen2008; @kenkel2013]. In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@oneil2014; @lemay2013].

## Conclusions

In this work, we bring reaction norms to the genomic era by characterizing the thermal reactionomes of the temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. We found that at least 10% of the transcriptome is thermally-responsive. For both species, more genes responded to cold than hot temperatures. In light of physiological data showing more variation in critical minimum than maximum temperatures, this suggests that these species and other ectotherms have less adaptive potential to hot than cold temperatures. Further, our results indicate that the two species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock and other protective molecules, while *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace. These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

# References
